Robustly decorrelating errors with mixed quantum gates

### 2019-09-27 decorrelation/mixed_unitaries.tex
Edits

CPTP maps possess a number of useful representations, including Kraus operators\cite{1983}, Choi matrices\cite{Choi1975}, and Jamiolkowsi states\cite{yczkowski2004}. But for the purposes of this article, the \emph{process matrix} representation will be particularly convenient\cite{OBrien2004}, and we shall denote the process matrix associated with a given CPTP map $\target$ with its corresponding calligraphic character, $\tarmat$. For a $d$-dimensional quantum state, the process matrix is a $d^2\times d^2$ matrix. A key feature of process matrices is that they are composable and act through the usual matrix multiplication on the vectorized quantum state:
\begin{align}
	\vectorize{\actual(\rho)}
		&= \actmat\cdot\vectorize{\rho} \\
		&= \errmat\cdot\tarmat\cdot\vectorize{\rho}
\end{align}
The $\mathsf{vec}$ operation is typically performed using a basis of matrix units, for which $\vectorize{\rho}$ would be the column vector obtained by stacking the columns of $\rho$. In this work, however, we shall use a basis of Pauli matrices, defining $P = \left\{I, \sigma_x, \sigma_y, \sigma_z\right\}^{\otimes n}$ as the collection of all $4^n$ $n$-qubit Pauli operators (including the identity).  In this basis,
\begin{equation}
  \vectorize{\rho}_i = \expect{P_i} = \tr\!\left(P_i\; \rho \right),
\end{equation}
and
\begin{equation}
	(\actmat)_{ij} = \frac{1}{d}\tr\!\left(P_i \;\actual\!\left(P_j\right) \right).
\end{equation}
Process matrices written in this Pauli basis are often referred to as \emph{Pauli transfer matrices} \cite{Chow2012}.  Error maps represented in this basis take the particularly nice form,
\begin{equation}\label{eq:process_matrix}
\errmat =
	\left(\begin{array}{c|cccc}
		1 &  & \vec{0}^T & \\ 
		\hline & &  &  \\
		\vec{m} &  & R &  \\
		 &  &  & 
	\end{array} 	
	\right)
\end{equation}
The top row of all trace-preserving (TP) error maps is fixed to $\{1,0,0,0,\cdots\}$ and the remainder of the first column, $\vec{m}$, describes any deviations from unitality, as might arise from amplitude damping \cite{preskill1997lecture}. If the error map is unitary, then the error is called \textit{coherent} and the unital submatrix $R$ is a rotation matrix.  Importantly, if $R$ is diagonal, then the error is Pauli stochastic, with each diagonal entry corresponding to the probability that its associated Pauli error occurs in each application of the gate. \kcy{We might need more here defining coherent noise. There is a lot of literature on this, most of it inconsistent. Try Joel's Estimating the Coherence of Noise (terrible title) and Gottesman's paper on Maximally Sensitive States.}

@@

\subsection{Mixed Quantum Gates}
\label{sec:mqg}
\noindent Suppose that we have access to an ensemble of distinct implementations, 
$\{\actual_i\}$, of a target gate, $\target$.
Each time the gate is to be applied to the system, we randomly select an implementation from this ensemble such that the probability of drawing $\actual_i$ is $\weight_i$ (and we ensure that $\sum_i \weight_i=1$). This procedure is operationally indistinguishable from always applying the effective channel, 
\begin{align}
	\actual_{\rm eff} &= \sum_i \weight_i \actual_i = \left(\sum_i \weight_i \error_i \right) \circ\target
\end{align}
We call such randomized quantum operations \emph{mixed quantum gates} or MQGs. Error metrics for these MQGs can be computed in terms of their effective error map, 
\begin{equation}
	\label{eq:effective_error}
	\error_{\rm eff} = \sum_i \weight_i \error_i.
\end{equation}
Two important error metrics are the average gate infidelity (AGI), $\AGI$, \kcy{We need a citation for this, and check if it's right! Gives weird answer if E = I. I feel like it should be $(d + \tr(\errmat))/(d+d^2)$. Maybe cite Nielsen's paper, or Horodecki**3.} and the diamond distance to the target (or simply, the \emph{diamond distance}), $\dnorm$ \cite{watrous2018theory}, defined as:
\begin{align}
	\label{eq:def_agi}
	&\AGI(\error) = \frac{d^2 - \tr{\errmat}}{d^2 + d},\\
	\label{eq:def_diamond}
	&\dnorm \left(\error\right)
		= \frac{1}{2} \sup_\rho \vert \vert (\ident_d\otimes \ident_d)(\rho) 
										  - (\error \otimes \ident_d)(\rho) \vert\vert_1,
\end{align}
where $d=2^n$ and $\ident_d$ is the $d$-dimensional identity operator.  In Eq.~\eqref{eq:def_agi}, we have written the AGI for an error map, $\error$, in terms of its associated Pauli transfer matrix, $\errmat$. If the error channel is purely stochastic, then $\dnorm(\error) = \AGI(\error)$, but if the error channel has a unitary component, then the diamond distance will generically be larger than the average gate infidelity \cite{1511.00727}. The diamond distance is subaddative \cite{watrous2018theory} under gate composition, so is particularly useful for constructing bounds on quantum circuit performance: the total variation distance between the outcome probabilities of a faulty and perfect quantum circuit is less than or equal to the sum of the diamond distances for gates that compose the circuit \cite{aharonov1998quantum}.

@@

\subsubsection{Sparsity Constraints}
\label{sec:sparsity}
\noindent
As a practical consideration, we would also like to regularize our objective function to promote sparse weightings. Control electronics often have a limited amount of waveform memory, and thus it is important to be able to construct MQGs as few constituent gates as possible. In many machine learning contexts, lasso regularization \cite{tibshirani1996regression} can be used to enforce sparsity in solutions. Rather than penalizing the $\ell_0$ norm of the solution, which is not convex, lasso regularization penalizes the $\ell_1$, which is the convex relaxation of the $\ell_0$ norm. However this is insufficient for our purposes, as we require $\weight$ to be a valid probability distribution, which constrains the $\ell_1$ norm to be equal to $1$. Conveniently, the problem of enforcing sparsity in such situations has been considered in \cite{NIPS2012_4504} and can be expressed via another convex program that extends (\ref{eq:robust_minimization}):

\kcy{Make the rest of this section consistent with the current notation.}
\begin{equation}\label{eq:minimization_regularization}
\begin{split}
&\underset{m\in[N]}{\textbf{minimize}}\\
&\ \ \bunderbrace{\textbf{minimize}: }{\weight_i\geq0, |\weight|_1=1,\\ t\geq0} ||{\mathbf{D}^{\ell}}\weight||_1 + t\\
&\ \ \textbf{subject to: } \weight_m > \frac{\lambda}{t}\\
&\ \ \phantom{\textbf{subject to: }} \forall n<\ell, ||\sum \weight_iD_i^n||=0
\end{split}
\end{equation} with $\lambda\geq0$. As with $\eta$ in the last subsection, the optimal value of $\lambda$ is problem specific, depending on $\mathbf{D}^\ell$ and how sparse the solution needs to be. In the following sections we discuss both experimental and numerical implementations of MQGs, leveraging these optimization techniques.

@@

% ------------------------------------------------------------------------------
% Subsection: Single-qubit example
% ------------------------------------------------------------------------------
\subsection{Single-qubit example} 
\label{sub:one_qubit}
We consider the following dimensionless model for a single tunable qubit: 
\begin{equation}\label{eq:1Qham}
  H(\delta, \epsilon, t) = \epsilon\sigma_z + (1 + \delta)(c_x(t)\sigma_x + c_y(t)\sigma_y)
\end{equation}
where $\epsilon$ corresponds to fluctuations in qubit frequency and $\delta$ corresponds to fluctuations in the control field.

To generate the initial controls, we use the GRAPE algorithm\cite{Khaneja2005} with N=25 steps and total evolution time of $\pi$ to generate 100 candidate controls. However, by imposing the sparisty constraint discussed in Section \ref{sec:sparsity}, we found MQGs consisting of just 10 controls. In our implementation of the GRAPE algorithm, we use the performance function presented in \cite{Khaneja2005}, and average over different values of $\delta$ and $\epsilon$ using Gaussian quadrature when computing the gradient, so that we find controls that are naturally robust. The standard deviations considered for all parameters in our numerical experiments were fixed to $\sigma=.001$. Finally, we assume that the errors on $\sigma_x$ and $\sigma_y$ are perfectly correlated, as in our experimental implementation. We note that advanced quantum control protocols may provide an even more principled approach to control synthesis. DMORPH \cite{dominy2008exploring}, for example, explores continuous families of controls on fidelity level sets, thereby enabling further optimizations against secondary criteria, such as the duration of the control pulse or robustness to drift.

@@

% ------------------------------------------------------------------------------
% Subsection: Single-qubit example
% ------------------------------------------------------------------------------
\subsection{Two-qubit example} 
\label{sub:two_qubit}
In our two-qubit example we consider the following model for two tunable qubits coupled by a resonant exchange interaction, similar to that in \cite{McKay2016}:
\begin{equation} \label{eq:2Qham}
\begin{split}
H(\vec{\delta}, \vec{\epsilon}, t) = &\sum_{j=1}^2(\epsilon_j\sigma_z^j + (1 + \delta_j)(c_x^j(t)\sigma_x^j + c_y^j(t)\sigma_y^j)) \\
&+ \frac{1}{10}(XX + YY)
\end{split}
\end{equation}

\amp{This section is still tough to read, but I'm not sure of a better way to handle it. The constituent controls are clearly contrived, so I think it best to be explicit about how they were constructed.}In this example it was infeasible to use GRAPE to return non-trivial solutions. Instead we manually selected piecewise constant echoing sequences with 500 steps and total evolution time of $\frac{5\pi}{2}$. In particular, we considered $RX(\pi)$, $RX(-\pi)$, $RY(\pi)$ and $RY(-\pi)$ bang-bang sequences \cite{bangbang}, consisting of all combinations of simultaneous $\pi$ pulses activated at multiples of $8$ steps from the beginning of the controls, and the same multiple of $8$ steps prior to the end of the controls. To give the control family a variety of RF errors, we also added uniformly distributed amplitude errors to each $\pi$ pulse, between $-.25$\% and $.25$\%.

>>>>

CPTP maps possess a number of useful representations, including Kraus operators\cite{1983}, Choi matrices\cite{Choi1975}, and Jamiolkowsi states\cite{yczkowski2004}. But for the purposes of this article, the \emph{process matrix} representation will be particularly convenient\cite{OBrien2004}, and we shall denote the process matrix associated with a given CPTP map $\target$ with its corresponding calligraphic character, $\tarmat$. For a $d$-dimensional quantum state, the process matrix is a $d^2\times d^2$ matrix. A key feature of process matrices is that they are composable and act through the usual matrix multiplication on the vectorized quantum state:
\begin{align}
	\vectorize{\actual(\rho)}
		&= \actmat\cdot\vectorize{\rho} \\
		&= \errmat\cdot\tarmat\cdot\vectorize{\rho}
\end{align}
The $\mathsf{vec}$ operation is typically performed using a basis of matrix units, for which $\vectorize{\rho}$ would be the column vector obtained by stacking the columns of $\rho$. In this work, however, we shall use a basis of Pauli matrices, defining $P = \left\{I, \sigma_x, \sigma_y, \sigma_z\right\}^{\otimes n}$ as the collection of all $4^n$ $n$-qubit Pauli operators (including the identity).  In this basis,
\begin{equation}
  \vectorize{\rho}_i = \expect{P_i} = \tr\!\left(P_i\; \rho \right),
\end{equation}
and
\begin{equation}
	(\actmat)_{ij} = \frac{1}{d}\tr\!\left(P_i \;\actual\!\left(P_j\right) \right).
\end{equation}
Process matrices written in this Pauli basis are often referred to as \emph{Pauli transfer matrices} \cite{Chow2012}.  Error maps represented in this basis take the particularly nice form,
\begin{equation}\label{eq:process_matrix}
\errmat =
	\left(\begin{array}{c|cccc}
		1 &  & \vec{0}^T & \\ 
		\hline & &  &  \\
		\vec{m} &  & R &  \\
		 &  &  & 
	\end{array} 	
	\right)
\end{equation}
The top row of all trace-preserving (TP) error maps is fixed to $\{1,0,0,0,\cdots\}$ and the remainder of the first column, $\vec{m}$, describes any deviations from unitality, as might arise from amplitude damping \cite{preskill1997lecture}. If the error map is unitary, then the error is called \textit{coherent} and the unital submatrix $R$ is a rotation matrix.  Importantly, if $R$ is diagonal, then the error is Pauli stochastic, with each diagonal entry corresponding to the probability that its associated Pauli error occurs in each application of the gate. \kcy{We might need more here defining coherent noise. There is a lot of literature on this, most of it inconsistent. Try Joel's Estimating the Coherence of Noise (terrible title) and Gottesman's paper on Maximally Sensitive States, and Arnaud's work on the Polar Decomposition.}

@@

\subsection{Mixed Quantum Gates}
\label{sec:mqg}
\noindent Suppose that we have access to an ensemble of distinct implementations, 
$\{\actual_i\}$, of a target gate, $\target$.
Each time the gate is to be applied to the system, we randomly select an implementation from this ensemble such that the probability of drawing $\actual_i$ is $\weight_i$ (and we ensure that $\sum_i \weight_i=1$). This procedure is operationally indistinguishable from always applying the effective channel, 
\begin{align}
	\actual_{\rm eff} &= \sum_i \weight_i \actual_i = \left(\sum_i \weight_i \error_i \right) \circ\target
\end{align}
We call such randomized quantum operations \emph{mixed quantum gates} or MQGs. Error metrics for these MQGs can be computed in terms of their effective error map, 
\begin{equation}
	\label{eq:effective_error}
	\error_{\rm eff} = \sum_i \weight_i \error_i.
\end{equation}
Two important error metrics are the average gate infidelity (AGI), $\AGI$, \kcy{We need a citation for this, and check if it's right! Gives weird answer if E = I. I feel like it should be $(d + \tr(\errmat))/(d+d^2)$. Maybe cite Nielsen's paper, or Horodecki**3, ** cite QUANTUM GATE FIDELITY IN TERMS OF CHOI MATRICES NATHANIEL JOHNSTON AND DAVID W. KRIBS
.} and the diamond distance to the target (or simply, the \emph{diamond distance}), $\dnorm$ \cite{watrous2018theory}, defined as:
\begin{align}
	\label{eq:def_agi}
	&\AGI(\error) = \frac{d^2 - \tr{\errmat}}{d^2 + d},\\
	\label{eq:def_diamond}
	&\dnorm \left(\error\right)
		= \frac{1}{2} \sup_\rho \vert \vert (\ident_d\otimes \ident_d)(\rho) 
										  - (\error \otimes \ident_d)(\rho) \vert\vert_1,
\end{align}
where $d=2^n$ and $\ident_d$ is the $d$-dimensional identity operator.  In Eq.~\eqref{eq:def_agi}, we have written the AGI for an error map, $\error$, in terms of its associated Pauli transfer matrix, $\errmat$. If the error channel is purely stochastic, then $\dnorm(\error) = \AGI(\error)$, but if the error channel has a unitary component, then the diamond distance will generically be larger than the average gate infidelity \cite{1511.00727}. The diamond distance is subaddative \cite{watrous2018theory} under gate composition, so is particularly useful for constructing bounds on quantum circuit performance: the total variation distance between the outcome probabilities of a faulty and perfect quantum circuit is less than or equal to the sum of the diamond distances for gates that compose the circuit \cite{aharonov1998quantum}.

@@

\subsubsection{Sparsity Constraints}
\label{sec:sparsity}
\noindent
As a practical consideration, we would also like to regularize our objective function to promote sparse weightings. Control electronics often have a limited amount of waveform memory, and thus it is important to be able to construct MQGs as few constituent gates as possible. In many machine learning contexts, lasso regularization \cite{tibshirani1996regression} can be used to enforce sparsity in solutions. Rather than penalizing the $\ell_0$ norm of the solution, which is not convex, lasso regularization penalizes the $\ell_1$, which is the convex relaxation of the $\ell_0$ norm. However this is insufficient for our purposes, as we require $\weight$ to be a valid probability distribution, which constrains the $\ell_1$ norm to be equal to $1$. Conveniently, the problem of enforcing sparsity in such situations has been considered in \cite{NIPS2012_4504} and can be expressed via another convex program that extends Eq.~\eqref{eq:minimization}:
\begin{equation}\label{eq:minimization_regularization}
\begin{split}
&\underset{m\in[1..N]}{\textbf{minimize:}}\\
&\ \ \bunderbrace{\textbf{\hspace{.2cm}minimize:\hspace{.2cm}} }{\hspace{.1cm}\weight_i\geq0, |\weight|_1=1, t\geq0} ||\mathbf{L}\cdot\vec\weight||_2 + t\\
&\ \ \hspace{.2cm}\textbf{subject to: } \weight_m > \frac{\lambda}{t}.
\end{split}
\end{equation}
In this program, $\lambda > 0$ is a tunable parameter that can be used to control the degree of sparsity in the solution.

In the following sections we discuss both experimental and numerical implementations of MQGs, leveraging these optimization techniques.

@@

% ------------------------------------------------------------------------------
% Subsection: Single-qubit example
% ------------------------------------------------------------------------------
\subsubsection{Single-qubit example} 
\label{sub:one_qubit}
We consider the following dimensionless model for a single tunable qubit: 
\begin{equation}\label{eq:1Qham}
  H(\delta, \epsilon, t) = \epsilon\sigma_z + (1 + \delta)(c_x(t)\sigma_x + c_y(t)\sigma_y)
\end{equation}
where $\epsilon$ corresponds to fluctuations in qubit frequency and $\delta$ corresponds to fluctuations in the control field.

To generate the initial controls, we use the GRAPE algorithm\cite{Khaneja2005} with N=25 steps and total evolution time of $\pi$ to generate 100 candidate controls. However, by imposing the sparisty constraint discussed in Section \ref{sec:sparsity}, we found MQGs consisting of just \kcy{10 controls}. In our implementation of the GRAPE algorithm, we use the performance function presented in \cite{Khaneja2005}, and average over different values of $\delta$ and $\epsilon$ using Gaussian quadrature when computing the gradient, so that we find controls that are naturally robust. The standard deviations considered for all parameters in our numerical experiments were fixed to $\sigma=.001$. Finally, we assume that the errors on $\sigma_x$ and $\sigma_y$ are perfectly correlated, as in our experimental implementation. We note that advanced quantum control protocols may provide an even more principled approach to control synthesis. DMORPH \cite{dominy2008exploring}, for example, explores continuous families of controls on fidelity level sets, thereby enabling further optimizations against secondary criteria, such as the duration of the control pulse or robustness to drift.

@@

% ------------------------------------------------------------------------------
% Subsection: Single-qubit example
% ------------------------------------------------------------------------------
\subsubsection{Two-qubit example} 
\label{sub:two_qubit}
In our two-qubit example we consider the following model for two tunable qubits coupled by a resonant exchange interaction, similar to that in \cite{McKay2016}:
\begin{equation} \label{eq:2Qham}
\begin{split}
H(\vec{\delta}, \vec{\epsilon}, t) = &\sum_{j=1}^2(\epsilon_j\sigma_z^j + (1 + \delta_j)(c_x^j(t)\sigma_x^j + c_y^j(t)\sigma_y^j)) \\
&+ \frac{1}{10}(XX + YY)
\end{split}
\end{equation}

\amp{This section is still tough to read, but I'm not sure of a better way to handle it. The constituent controls are clearly contrived, so I think it best to be explicit about how they were constructed.}In this example it was infeasible to use GRAPE to return non-trivial solutions (always got the same result). Instead we manually selected piecewise constant echoing sequences with 500 steps and total evolution time of $\frac{5\pi}{2}$. In particular, we considered $RX(\pi)$, $RX(-\pi)$, $RY(\pi)$ and $RY(-\pi)$ bang-bang sequences \cite{bangbang}, consisting of all combinations of simultaneous $\pi$ pulses activated at multiples of $8$ steps from the beginning of the controls, and the same multiple of $8$ steps prior to the end of the controls. To give the control family a variety of RF errors, we also added uniformly distributed amplitude errors to each $\pi$ pulse, between $-.25$\% and $.25$\%.
